---
title: "P8105 Homework 2"
author: "Zheming Bao"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(readxl)
```

# Problem 1

```{r subway_import}
subway =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, vending, entrance_type, ada
  ) |>
  mutate(entry = entry == "YES")

dim(subway)
```

The cleaned data contain `r nrow(subway)` entrances/exits and
`r ncol(subway)` variables: line, station name and coordinates, routes,
entry, vending, entrance type, and ADA status. I cleaned column names,
selected these variables, and converted entry to logical. The data are
not tidy because routes occupy multiple columns.

```{r subway_summary}
n_stations =
  subway |>
  distinct(station_name, line) |>
  nrow()

n_ada_stations =
  subway |>
  filter(ada) |>
  distinct(station_name, line) |>
  nrow()

no_vending =
  subway |>
  filter(vending == "NO") |>
  summarise(
    entrances_exits = n(),
    allow_entry = sum(entry),
    proportion_allow_entry = mean(entry)
  )

n_stations
n_ada_stations
no_vending
```

There are **`r n_stations` stations** (identified by name and line),
including **`r n_ada_stations` ADA-compliant stations**. Without vending,
`r (no_vending |> pull(allow_entry))` of `r (no_vending |> pull(entrances_exits))` entrances/exits
allow entry (**`r round(100 * (no_vending |> pull(proportion_allow_entry)), 2)`%**).

```{r subway_routes}
subway_long =
  subway |>
  mutate(across(starts_with("route"), as.character)) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(route_number = as.integer(route_number))

a_stations =
  subway_long |>
  filter(route_name == "A") |>
  distinct(station_name, line, ada)

n_a_stations =
  a_stations |>
  distinct(station_name, line) |>
  nrow()

n_a_ada_stations =
  a_stations |>
  filter(ada) |>
  distinct(station_name, line) |>
  nrow()

n_a_stations
n_a_ada_stations
```

I used pivot_longer() to put route numbers and names in two columns,
dropping missing routes.
**`r n_a_stations` stations serve the A train; `r n_a_ada_stations` are
ADA compliant.**

# Problem 2

```{r trash_wheel_import}
trash_file = "data/202610 Trash Wheel Collection Data.xlsx"

mr_trash =
  read_excel(
    trash_file,
    sheet = "Mr. Trash Wheel", range = "A2:N742",
    na = "dive"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

professor_trash =
  read_excel(
    trash_file,
    sheet = "Professor Trash Wheel", range = "A2:M139",
    na = "dive"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda_trash =
  read_excel(
    trash_file,
    sheet = "Gwynnda the Good Wheel of the W", range = "A2:L394",
    na = "dive"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    date = as.Date(date),
    trash_wheel = "Gwynnda"
  )

trash_wheel =
  bind_rows(mr_trash, professor_trash, gwynnda_trash) |>
  relocate(trash_wheel, dumpster)

trash_wheel |>
  count(trash_wheel, name = "observations")
```

I cleaned names, removed rows without dumpster numbers, and treated
"dive" as missing. I converted year and date types, rounded sports balls
to integers, and combined the tables using a wheel identifier.

```{r trash_wheel_summary}
professor_weight =
  trash_wheel |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarise(total_weight_tons = sum(weight_tons, na.rm = TRUE))

gwynnda_june_2022 =
  trash_wheel |>
  filter(trash_wheel == "Gwynnda", year == 2022, month == "June") |>
  summarise(total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE))

professor_weight
gwynnda_june_2022
```

The combined data have **`r nrow(trash_wheel)` dumpster records** and
**`r ncol(trash_wheel)` variables**, including wheel, date, weight (tons),
volume (cubic yards), and litter counts. Professor Trash Wheel collected
**`r round((professor_weight |> pull(total_weight_tons)), 2)` tons** in total.
Gwynnda collected **`r (gwynnda_june_2022 |> pull(total_cigarette_butts))`
cigarette butts in June 2022**.

# Problem 3

```{r baseline_import}
baseline_raw =
  read_csv(
    "data/MCI_baseline.csv", skip = 1,
    na = c("", "NA", "."), show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  rename(baseline_age = current_age) |>
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    apoe4 = factor(apoe4, levels = c(0, 1),
                   labels = c("Non-carrier", "Carrier"))
  )

baseline =
  baseline_raw |>
  filter(is.na(age_at_onset) | age_at_onset > baseline_age)

baseline_summary =
  baseline |>
  summarise(
    eligible = n(),
    developed_mci = sum(!is.na(age_at_onset)),
    mean_baseline_age = mean(baseline_age)
  )

female_carriers =
  baseline |>
  filter(sex == "Female") |>
  summarise(proportion = mean(apoe4 == "Carrier"))

baseline_summary
female_carriers
```

I skipped the description row, treated dots as missing, cleaned names,
and labeled sex and APOE4 status as factors. Of **`r nrow(baseline_raw)`
recruited participants**, `r nrow(baseline_raw) - nrow(baseline)` had MCI at
or before baseline and were excluded. The remaining **`r nrow(baseline)`
participants** include **`r (baseline_summary |> pull(developed_mci))` who developed
MCI**. Their mean baseline age is **`r round((baseline_summary |> pull(mean_baseline_age)), 2)`
years**, and **`r round(100 * (female_carriers |> pull(proportion)), 2)`% of women**
are APOE4 carriers. Each row contains one participant's demographics and
MCI onset age; missing onset age indicates no MCI during follow-up.

```{r amyloid_import}
amyloid_wide =
  read_csv(
    "data/mci_amyloid.csv", skip = 1,
    na = c("", "NA", "."), show_col_types = FALSE
  ) |>
  janitor::clean_names() |>
  rename(id = study_id, time_0 = baseline)

amyloid =
  amyloid_wide |>
  pivot_longer(
    cols = starts_with("time_"),
    names_to = "year", names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |>
  mutate(year = as.integer(year))

amyloid |>
  summarise(
    participants = n_distinct(id),
    rows = n(),
    missing_ratios = sum(is.na(amyloid_ratio))
  )
```

I skipped the description row and put visits (years 0, 2, 4, 6, and 8)
in long format. The tidy data contain
**`r n_distinct((amyloid |> pull(id)))` participants** and **`r nrow(amyloid)`
participant-visit rows**. The **`r sum(is.na((amyloid |> pull(amyloid_ratio))))` missing
ratios** are retained as NA.

```{r compare_participants}
baseline_only =
  anti_join(baseline, amyloid_wide, by = "id")
amyloid_only =
  anti_join(amyloid_wide, baseline, by = "id")

baseline_only |> select(id)
amyloid_only |> select(id)

# Compare IDs before exclusions.
raw_baseline_only =
  anti_join(baseline_raw, amyloid_wide, by = "id")
raw_amyloid_only =
  anti_join(amyloid_wide, baseline_raw, by = "id")
```

After applying the inclusion criterion, **`r nrow(baseline_only)`
participants appear only in baseline** (IDs:
`r paste((baseline_only |> pull(id)), collapse = ", ")`), and **`r nrow(amyloid_only)`
appear only in amyloid** (IDs: `r paste((amyloid_only |> pull(id)), collapse = ", ")`).
Before exclusion, the source files have `r nrow(raw_baseline_only)`
baseline-only and `r nrow(raw_amyloid_only)` amyloid-only IDs. The other
`r nrow(amyloid_only) - nrow(raw_amyloid_only)` amyloid-only participants
were excluded for MCI at baseline.

```{r merge_export}
combined_mci =
  inner_join(baseline, amyloid, by = "id") |>
  arrange(id, year)

write_csv(combined_mci, "data/combined_mci.csv")

combined_mci |>
  summarise(
    participants = n_distinct(id),
    rows = n(),
    missing_ratios = sum(is.na(amyloid_ratio))
  )
```

The inner join retains **`r n_distinct((combined_mci |> pull(id)))` eligible participants
present in both datasets**, with **`r nrow(combined_mci)` rows and
`r ncol(combined_mci)` columns**. Each row combines demographics with one
visit's amyloid ratio; `r sum(is.na((combined_mci |> pull(amyloid_ratio))))` ratios are
missing. The result is saved as `data/combined_mci.csv`.
